---
title: "01 - Data Exploration: GSE25066"
author: "Maria Szczerbinska"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

# Overview

This notebook performs initial exploration of GSE25066 - a microarray 
dataset of 508 primary breast tumours from tamoxifen-treated patients, 
with treatment response annotation (pathological complete response, pCR, 
vs. residual disease, RD).

**Biological question:** Are genes involved in the FGFR2-ERα signalling 
axis differentially expressed between tamoxifen-sensitive and 
tamoxifen-resistant tumours?

**Genes of interest:** FGFR2, ESR1, PGR, IRS1, BCL2L1

**Data source:** NCBI GEO, accession GSE25066  
**Platform:** Affymetrix HGU133A microarray  
**Analysis method:** limma (linear models for microarray data)

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
# Install if missing: BiocManager::install("limma"); BiocManager::install("GEOquery")

library(GEOquery)   # load GEO data
library(limma)      # differential expression for microarray
library(ggplot2)    # visualisation
library(dplyr)      # data manipulation

# Set ggplot2 theme for all plots
theme_set(theme_bw(base_size = 13))
```

---

# 2. Load data

```{r load-data, message=FALSE, warning=FALSE}
# Load the downloaded series matrix file
gse <- getGEO(
  "GSE25066",
  destdir   = "../data/raw",
  GSEMatrix = TRUE
)

# Extract ExpressionSet object
eset <- gse[[1]]

# Basic dimensions
cat("Number of samples: ", ncol(eset),  "\n")
cat("Number of probes:  ", nrow(eset),  "\n")
```

---

# 3. Explore clinical (phenotype) data

```{r pheno-explore}
# Extract phenotype data (clinical annotations)
pheno <- pData(eset)

# Preview column names - we need to find tamoxifen response annotation
colnames(pheno)
```

```{r pheno-response}
# GSE25066 contains treatment response as pCR (complete response)
# vs RD (residual disease) - key variable for our analysis

# Preview first few rows of relevant columns
head(pheno[, c(
  "title",
  "characteristics_ch1",
  "characteristics_ch1.1",
  "characteristics_ch1.2"
)])
```

```{r response-counts}
# Find and tabulate tamoxifen response
# Note: column name may vary - adjust based on colnames() output above

# Common column name in GSE25066:
response_col <- "pathologic_response_pcr_rd:ch1"

response_table <- table(pheno[[response_col]])
print(response_table)

# Plot sample distribution by response
ggplot(pheno, aes(x = pheno[[response_col]])) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Sample distribution by tamoxifen response",
    x     = "Treatment response",
    y     = "Number of samples",
    caption = "GSE25066 | n = 508"
  )

ggsave(
  "../results/figures/01_sample_distribution.png",
  width = 6, height = 4, dpi = 150
)
```

---

# 4. Quality control

```{r qc-boxplot, fig.width=12, fig.height=5}
# Expression value distribution across samples
# A well-normalised dataset should show similar distributions per sample

# Take a random subset of 40 samples for readability
set.seed(42)
sample_idx <- sample(ncol(eset), 40)

boxplot(
  exprs(eset)[, sample_idx],
  outline = FALSE,
  las     = 2,
  cex.axis = 0.5,
  col      = "steelblue",
  main     = "Expression distribution (random subset of 40 samples)",
  ylab     = "Expression value (log2)"
)
```

```{r qc-density}
# Density plot of expression values - all samples overlaid
# All lines should roughly overlap if data is well-normalised

expr_df <- as.data.frame(exprs(eset)[, sample_idx]) |>
  tidyr::pivot_longer(
    cols      = everything(),
    names_to  = "sample",
    values_to = "expression"
  )

ggplot(expr_df, aes(x = expression, group = sample)) +
  geom_density(alpha = 0.3, colour = "steelblue", linewidth = 0.3) +
  labs(
    title   = "Expression density (random subset of 40 samples)",
    x       = "Expression value (log2)",
    y       = "Density",
    caption = "GSE25066"
  )

ggsave(
  "../results/figures/01_expression_density.png",
  width = 7, height = 4, dpi = 150
)
```

---

# 5. PCA - overview of sample structure

```{r pca}
# PCA helps detect batch effects and outliers
# and shows whether samples cluster by biological group

# Transpose: PCA expects samples as rows
pca_result <- prcomp(
  t(exprs(eset)),
  center = TRUE,
  scale. = FALSE
)

# Variance explained by each PC
var_explained <- round(
  summary(pca_result)$importance[2, 1:5] * 100, 1
)

# Build PCA dataframe for plotting
pca_df <- data.frame(
  PC1      = pca_result$x[, 1],
  PC2      = pca_result$x[, 2],
  response = pheno[[response_col]]
)

ggplot(pca_df, aes(x = PC1, y = PC2, colour = response)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_colour_manual(values = c("pCR" = "#7BA7E0", "RD" = "#E07B7B", "NA" = "grey70")) +
  labs(
    title   = "PCA of GSE25066 samples",
    x       = paste0("PC1 (", var_explained[1], "% variance)"),
    y       = paste0("PC2 (", var_explained[2], "% variance)"),
    colour  = "Response",
    caption = "GSE25066 | n = 508"
  )

ggsave(
  "../results/figures/01_pca.png",
  width = 7, height = 5, dpi = 150
)
```

---

# 6. Genes of interest - quick look

```{r genes-of-interest}
# Check expression of key FGFR2-ERα pathway genes
# Note: microarray uses probe IDs, not gene symbols
# We need to map probe IDs to gene symbols via feature data

# Extract feature data (probe - gene symbol mapping)
fdata <- fData(eset)
head(fdata[, c("ID", "Gene Symbol", "Gene Title")])
```

```{r gene-lookup}
# Genes of interest
genes_of_interest <- c("FGFR2", "ESR1", "PGR", "IRS1", "BCL2L1")

# Find probes corresponding to these genes
probes_of_interest <- fdata[
  fdata[["Gene Symbol"]] %in% genes_of_interest,
  c("ID", "Gene Symbol")
]

print(probes_of_interest)
cat("\nFound", nrow(probes_of_interest), "probes for", 
    length(genes_of_interest), "genes of interest\n")
```

---

# 7. Session info

```{r session-info}
sessionInfo()
```

---

# Next steps

- `02_differential_expression.Rmd` - limma analysis: pCR vs RD
- `03_pathway_analysis.Rmd` - FGFR2-ERα pathway visualisation
- `04_survival_analysis.Rmd` - correlation with clinical outcomes
