---
title: "02 - Differential Expression Analysis: pCR vs RD"
author: "Maria Szczerbinska"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

# Overview

This notebook performs differential expression analysis between tamoxifen-sensitive 
(pCR) and tamoxifen-resistant (RD) tumours using limma.

We focus on the FGFR2-ERα signalling axis genes: FGFR2, ESR1, PGR, IRS1, BCL2L1.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(GEOquery)
library(limma)
library(ggplot2)
library(dplyr)

theme_set(theme_bw(base_size = 13))
```

---

# 2. Load data

```{r load-data, message=FALSE, warning=FALSE}
gse  <- getGEO("GSE25066", destdir = "../data/raw", GSEMatrix = TRUE)
eset <- gse[[1]]

# Extract expression matrix and phenotype data
expr  <- exprs(eset)
pheno <- pData(eset)
fdata <- fData(eset)

cat("Samples:", ncol(expr), "\n")
cat("Probes: ", nrow(expr), "\n")
```

---

# 3. Prepare groups

```{r prepare-groups}
# Extract response column and keep only pCR and RD samples
response <- pheno[["pathologic_response_pcr_rd:ch1"]]

# Remove NA samples
keep     <- !is.na(response) & response %in% c("pCR", "RD")
expr_sub  <- expr[, keep]
response_sub <- factor(response[keep], levels = c("RD", "pCR"))

cat("Samples after removing NA:\n")
print(table(response_sub))
```

---

# 4. Probe selection - one probe per gene

```{r probe-selection}
# For genes with multiple probes, keep the probe with highest mean expression
# This is a standard approach for Affymetrix data

genes_of_interest <- c("FGFR2", "ESR1", "PGR", "IRS1", "BCL2L1")

# Find all probes for genes of interest
probes <- fdata[fdata[["Gene Symbol"]] %in% genes_of_interest,
                c("ID", "Gene Symbol")]

# Calculate mean expression per probe
probes$mean_expr <- rowMeans(expr_sub[probes$ID, ])

# Select probe with highest mean expression per gene
best_probes <- probes %>%
  group_by(`Gene Symbol`) %>%
  slice_max(mean_expr, n = 1) %>%
  ungroup()

print(best_probes[, c("Gene Symbol", "ID", "mean_expr")])
```

---

# 5. Differential expression with limma

```{r limma-full}
# Run limma on full expression matrix
# This gives us genome-wide results

# Design matrix: intercept + group effect
design <- model.matrix(~ response_sub)

# Fit linear model
fit  <- lmFit(expr_sub, design)
fit  <- eBayes(fit)

# Extract all results
results_all <- topTable(fit, coef = 2, number = Inf, sort.by = "P")

# Add gene symbols
results_all$probe_id    <- rownames(results_all)
results_all$gene_symbol <- fdata[results_all$probe_id, "Gene Symbol"]

cat("Total probes tested:", nrow(results_all), "\n")
cat("Significant (adj.P < 0.05):", sum(results_all$adj.P.Val < 0.05), "\n")
cat("Significant (adj.P < 0.05 & |logFC| > 1):",
    sum(results_all$adj.P.Val < 0.05 & abs(results_all$logFC) > 1), "\n")
```

---

# 6. Results for genes of interest

```{r genes-results}
# Extract limma results for our FGFR2-ERα pathway genes
genes_results <- results_all[results_all$probe_id %in% best_probes$ID, ]
genes_results$gene_symbol <- fdata[genes_results$probe_id, "Gene Symbol"]

# Select relevant columns
genes_results <- genes_results[, c("gene_symbol", "logFC", "AveExpr",
                                    "t", "P.Value", "adj.P.Val")]

# Sort by p-value
genes_results <- genes_results[order(genes_results$P.Value), ]

print(genes_results)
```

---

# 7. Volcano plot

```{r volcano, fig.width=8, fig.height=6}
# Classify probes for colouring
results_all$status <- "Not significant"
results_all$status[results_all$adj.P.Val < 0.05 & results_all$logFC >  1] <- "Up in pCR"
results_all$status[results_all$adj.P.Val < 0.05 & results_all$logFC < -1] <- "Up in RD"

# Label our genes of interest
results_all$label <- ifelse(
  results_all$probe_id %in% best_probes$ID,
  results_all$gene_symbol,
  NA
)

ggplot(results_all, aes(x = logFC, y = -log10(P.Value), colour = status)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_point(
    data = results_all[!is.na(results_all$label), ],
    size = 3
  ) +
  ggrepel::geom_text_repel(
    aes(label = label),
    size = 3.5,
    max.overlaps = 20
  ) +
  scale_colour_manual(values = c(
    "Not significant" = "grey70",
    "Up in pCR"       = "#7BA7E0",
    "Up in RD"        = "#E07B7B"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  labs(
    title   = "Differential expression: pCR vs RD",
    x       = "log2 Fold Change (pCR vs RD)",
    y       = "-log10(p-value)",
    colour  = "Status",
    caption = "GSE25066 | limma | adj.P < 0.05 & |logFC| > 1"
  )

ggsave(
  "../results/figures/02_volcano_plot.png",
  width = 8, height = 6, dpi = 150
)
```

---

# 8. Heatmap - top 50 genes + genes of interest

```{r heatmap, fig.width=8, fig.height=10}
# Select top 50 DE genes + our genes of interest
top50_probes <- head(rownames(results_all[results_all$adj.P.Val < 0.05, ]), 50)
probes_to_plot <- unique(c(top50_probes, best_probes$ID))

# Expression matrix for heatmap
expr_heatmap <- expr_sub[probes_to_plot, ]

# Replace probe IDs with gene symbols where available
gene_labels <- fdata[rownames(expr_heatmap), "Gene Symbol"]
gene_labels[gene_labels == ""] <- rownames(expr_heatmap)[gene_labels == ""]
rownames(expr_heatmap) <- gene_labels

# Annotation for columns (samples)
col_annotation <- data.frame(
  Response = response_sub,
  row.names = colnames(expr_heatmap)
)

ann_colours <- list(Response = c("pCR" = "#7BA7E0", "RD" = "#E07B7B"))

# Generate heatmap object
p <- pheatmap::pheatmap(
  expr_heatmap,
  annotation_col    = col_annotation,
  annotation_colors = ann_colours,
  show_colnames     = FALSE,
  scale             = "row",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color             = colorRampPalette(c("#7BA7E0", "white", "#E07B7B"))(100),
  main              = "Top DE genes + FGFR2-ERa pathway genes\n(pCR vs RD, GSE25066)",
  silent            = TRUE
)

# Display in notebook
print(p)

# Save to file separately
ggsave(
  "../results/figures/02_heatmap.png",
  plot   = p,
  width  = 8,
  height = 10,
  dpi    = 150
)
```

---

# 9. Session info

```{r session-info}
sessionInfo()
```

---

# Next steps

- `03_survival_analysis.Rmd` - Kaplan-Meier curves for top DE genes