---
title: "03 - Survival Analysis: FGFR2-ERa Pathway Genes"
author: "Maria Szczerbinska"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

# Overview

This notebook investigates whether expression of FGFR2-ERa pathway genes
(ESR1, PGR, FGFR2, IRS1, BCL2L1) correlates with distant relapse-free
survival (DRFS) in the GSE25066 cohort.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(GEOquery)
library(survival)
library(survminer)
library(dplyr)
library(ggplot2)

theme_set(theme_bw(base_size = 13))
```

---

# 2. Load data

```{r load-data, message=FALSE, warning=FALSE}
gse  <- getGEO("GSE25066", destdir = "../data/raw", GSEMatrix = TRUE)
eset <- gse[[1]]

expr  <- exprs(eset)
pheno <- pData(eset)
fdata <- fData(eset)
```

---

# 3. Prepare survival data

```{r prepare-survival}
# Extract survival variables
time_col  <- "drfs_even_time_years:ch1"
event_col <- "drfs_1_event_0_censored:ch1"

surv_data <- data.frame(
  sample_id = rownames(pheno),
  time      = as.numeric(pheno[[time_col]]),
  event     = as.numeric(pheno[[event_col]])
)

# Remove samples with missing survival data
surv_data <- surv_data[!is.na(surv_data$time) & !is.na(surv_data$event), ]

cat("Samples with survival data:", nrow(surv_data), "\n")
cat("Events (relapses):", sum(surv_data$event == 1), "\n")
cat("Censored:", sum(surv_data$event == 0), "\n")
```

---

# 4. Select probes for genes of interest

```{r probe-selection}
genes_of_interest <- c("FGFR2", "ESR1", "PGR", "IRS1", "BCL2L1")

probes <- fdata[fdata[["Gene Symbol"]] %in% genes_of_interest,
                c("ID", "Gene Symbol")]

probes$mean_expr <- rowMeans(expr[probes$ID, ])

best_probes <- probes %>%
  group_by(`Gene Symbol`) %>%
  slice_max(mean_expr, n = 1) %>%
  ungroup()

print(best_probes[, c("Gene Symbol", "ID")])
```

---

# 5. Build survival dataset with gene expression

```{r build-dataset}
# Extract expression for genes of interest, samples with survival data only
expr_genes <- t(expr[best_probes$ID, surv_data$sample_id])
colnames(expr_genes) <- best_probes$`Gene Symbol`

surv_full <- cbind(surv_data, expr_genes)

head(surv_full)
```

---

# 6. Kaplan-Meier curves - high vs low expression

For each gene, samples are split into "high" and "low" expression groups
using the median as cutoff.

```{r km-esr1, fig.width=7, fig.height=6}
# ESR1
surv_full$ESR1_group <- ifelse(
  surv_full$ESR1 > median(surv_full$ESR1), "High", "Low"
)

fit_esr1 <- survfit(
  Surv(time, event) ~ ESR1_group,
  data = surv_full
)

ggsurvplot(
  fit_esr1,
  data          = surv_full,
  pval          = TRUE,
  risk.table    = TRUE,
  palette       = c("#E07B7B", "#7BA7E0"),
  title         = "DRFS by ESR1 expression",
  xlab          = "Time (years)",
  ylab          = "Distant relapse-free survival",
  legend.title  = "ESR1 expression",
  legend.labs   = c("High", "Low")
)
```

```{r km-pgr, fig.width=7, fig.height=6}
# PGR
surv_full$PGR_group <- ifelse(
  surv_full$PGR > median(surv_full$PGR), "High", "Low"
)

fit_pgr <- survfit(
  Surv(time, event) ~ PGR_group,
  data = surv_full
)

ggsurvplot(
  fit_pgr,
  data          = surv_full,
  pval          = TRUE,
  risk.table    = TRUE,
  palette       = c("#E07B7B", "#7BA7E0"),
  title         = "DRFS by PGR expression",
  xlab          = "Time (years)",
  ylab          = "Distant relapse-free survival",
  legend.title  = "PGR expression",
  legend.labs   = c("High", "Low")
)
```

```{r km-fgfr2, fig.width=7, fig.height=6}
# FGFR2
surv_full$FGFR2_group <- ifelse(
  surv_full$FGFR2 > median(surv_full$FGFR2), "High", "Low"
)

fit_fgfr2 <- survfit(
  Surv(time, event) ~ FGFR2_group,
  data = surv_full
)

ggsurvplot(
  fit_fgfr2,
  data          = surv_full,
  pval          = TRUE,
  risk.table    = TRUE,
  palette       = c("#E07B7B", "#7BA7E0"),
  title         = "DRFS by FGFR2 expression",
  xlab          = "Time (years)",
  ylab          = "Distant relapse-free survival",
  legend.title  = "FGFR2 expression",
  legend.labs   = c("High", "Low")
)
```

---

# 7. Summary table - log-rank test results

```{r summary-table}
genes_to_test <- c("ESR1", "PGR", "FGFR2", "IRS1", "BCL2L1")

logrank_results <- lapply(genes_to_test, function(gene) {
  group_var <- ifelse(
    surv_full[[gene]] > median(surv_full[[gene]]), "High", "Low"
  )
  fit <- survdiff(Surv(time, event) ~ group_var, data = surv_full)
  p_value <- 1 - pchisq(fit$chisq, df = 1)
  data.frame(gene = gene, p_value = round(p_value, 4))
})

logrank_summary <- do.call(rbind, logrank_results)
logrank_summary <- logrank_summary[order(logrank_summary$p_value), ]

print(logrank_summary)
```

---

# 8. Session info

```{r session-info}
sessionInfo()
```